Estimation of Intracortical Connectivity in a Dynamical Neural Field Model

### 2009-12-04 ltx/BrainIDE.tex
added MATLAB code of model

\section{State-Space Model}
We begin the derivation of the state-space model with the discrete time, continuous space IDE neural field model
\begin{equation}\label{SystemEq}
{v_{t + 1}}\left( {\bf{r}} \right) = \int_\Omega  {k\left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{v_t}\left( {{\bf{r'}}} \right)} \right)d{\bf{r'}}}  - \alpha {v_t}\left( {\bf{r}} \right) + e_t\left( {\mathbf{r}} \right),
where $e_t(r)\sim \mathcal{GP}(0,R)$. The observation equation for the model is
\begin{equation}\label{ObservationEquation}
    y_t( k ) = \int_\Omega  m( k,\mathbf{r} )v_t( \mathbf{r}\rm ) d\mathbf{r}  + \varepsilon ( k ),
where $\varepsilon( k )\sim\mathcal{N}(0,\Sigma)$, $k$ indexes the iEEG electrodes, and $m(k,r)$ is the observation kernel relating the neural field to EEG measurements accounting for electrodes characteristics.
To infer the state from observations we can write the likelihood function
\begin{equation}\label{Likelihood}
p\left( {{y_{1,...,T}}|{v_{1,...,T}}} \right) = \prod\limits_{t = 1}^T {p\left( {{y_t}|{v_t}} \right)} \prod\limits_{t = 2}^T {p\left( {{v_{t + 1}}|{v_t}} \right)} p\left( {{v_0}} \right)
where
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  p\left( {{y_t}|{v_t}} \right) &=& \mathcal{N}\left( {\int_\Omega  {m\left( {k,r} \right)v_t( r )dr} ,R} \right) \\
  p\left( {{v_{t + 1}}|{v_t}} \right) &=& ??? \\
  p\left( {{v_0}} \right) &=& \mathcal{GP}\left( {{\mu _0},{\pi _0}} \right).
\end{eqnarray}
We can not calculate the posterior in this Bayesian model for some reason (Mike and Ken can ellaborate), so we need to decompose the field into weighted basis functions. Therefore we define
\begin{equation}\label{DefFieldDecomp}
{v_t}\left( {\bf{r}} \right) = {\phi ^T}\left( {\bf{r}} \right){x_t}
\begin{equation}\label{DefKernelDecomp}
k( \mathbf{r} , \mathbf{r'} ) = \theta^T(\mathbf{r})\psi(\mathbf{r} , \mathbf{r'} )
Will will assume we know the parametric form of the basis functions, where the parameter $\theta$ is unknown. Each connectivity basis function can be considered a layer in the WC model. Making substitutions we get
\begin{equation}\label{DecompModel}
{\phi ^T}\left( \mathbf{r} \right){x_{t+1}} = \theta^T(\mathbf{r})\int_\Omega  {\psi \left( {{\bf{r}} , {\mathbf{r'}}} \right)f\left( {{\phi ^T}\left( {{\mathbf{r'}}} \right){x_t}} \right)d{\mathbf{r'}}}  - \alpha\phi^T(r)x_t + {\varepsilon _t}\left( {\mathbf{r}} \right).
To isolate the state we cross multiply by $\phi(r)$ and integrate over $\bf{r}$
\begin{equation}\label{Poo}
\int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}{x_{t+1}}}  = \int_\Omega  {\phi ( \mathbf{r} )\theta^T(\mathbf{r})\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}}  - \alpha\int_\Omega \phi(r)\phi^T(r)d\bf{r} x_t  + \int_\Omega  {\phi \left( {\bf{r}} \right){\varepsilon _t}\left( {\bf{r}} \right)d{\bf{r}}}.
Now we define
\begin{equation}\label{DefGamma}
\Gamma  = \int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}}.
Substituting equation~\ref{DefGamma} into \ref{Poo} and rearranging gives
\begin{equation}\label{SS Model}
{x_{t+1}} = {\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)\theta^T(\mathbf{r})\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}}-\alpha x_t + {\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right){\varepsilon _t}\left( {\bf{r}} \right)d{\bf{r}}}.
Under the assumption of spatial homogeneity we can write
\begin{equation}\label{Homogeneous SS Model}
{x_{t+1}} = \theta^T{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)\int_\Omega  {\psi \left( {{\bf{r}} - {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}}-\alpha x_t + {\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right){\varepsilon _t}\left( {\bf{r}} \right)d{\bf{r}}}.
This provides the final state-space model
\begin{equation}\label{AbbrevSSModel}
x_{t + 1} = \theta^T q(x) -\alpha x + \Gamma ^{ - 1}\int_\Omega  {\phi ( \mathbf{r} ){\varepsilon _t}( \mathbf{r} )d\mathbf{r}}
with observation equation
\begin{equation}\label{ObservationEquation}
    y_t( k ) = \int_\Omega  m( k,\mathbf{r} )\phi^T( \mathbf{r} ) d\mathbf{r}x_t  + \varepsilon ( k ),

\section{Estimation of the Nonlinear Homogeneous IDE Neural Field Model}
Based on the other papers. Although it appears to be a job for a nonlinear versions of the Kalman filter.

>>>>

\section{State-Space Model}
We begin the derivation of the state-space model with the discrete time, continuous space IDE neural field model
\begin{equation}\label{SystemEq}
{v_{t + 1}}\left( {\bf{r}} \right) = \int_\Omega  {k\left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{v_t}\left( {{\bf{r'}}} \right)} \right)d{\bf{r'}}}  - \alpha {v_t}\left( {\bf{r}} \right) + e_t\left( {\mathbf{r}} \right),
\end{equation}
where $e_t(r)\sim \mathcal{GP}(0,R)$. The observation equation for the model is
\begin{equation}\label{ObservationEquation}
    y_t( k ) = \int_\Omega  m( k,\mathbf{r} )v_t( \mathbf{r}\rm ) d\mathbf{r}  + \varepsilon ( k ),
\end{equation}
where $\varepsilon( k )\sim\mathcal{N}(0,\Sigma)$, $k$ indexes the iEEG electrodes, and $m(k,r)$ is the observation kernel relating the neural field to EEG measurements accounting for electrodes characteristics.
To infer the state from observations we can write the likelihood function
\begin{equation}\label{Likelihood}
p\left( {{y_{1,...,T}}|{v_{1,...,T}}} \right) = \prod\limits_{t = 1}^T {p\left( {{y_t}|{v_t}} \right)} \prod\limits_{t = 2}^T {p\left( {{v_{t + 1}}|{v_t}} \right)} p\left( {{v_0}} \right)
\end{equation}
where
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  p\left( {{y_t}|{v_t}} \right) &=& \mathcal{N}\left( {\int_\Omega  {m\left( {k,r} \right)v_t( r )dr} ,R} \right) \\
  p\left( {{v_{t + 1}}|{v_t}} \right) &=& ??? \\
  p\left( {{v_0}} \right) &=& \mathcal{GP}\left( {{\mu _0},{\pi _0}} \right).
\end{eqnarray}
We can not calculate the posterior in this Bayesian model for some reason (Mike and Ken can ellaborate), so we need to decompose the field into weighted basis functions.

In order to implement standard estimation techniques we propose a decomposition of the The field using a set of Gaussian basis functions. Decomposition allows a continuous field to be represented by a finite dimensional state vector. This allows for the application of standard nonlinear, non-Gaussian state estimation methods such as sequential Monte-Carlo techniques like particle filtering. The field decomposition is described by
\begin{equation}\label{DefFieldDecomp}
{v_t}\left( {\bf{r}} \right) = {\phi ^T}\left( {\bf{r}} \right){x_t}.
\end{equation}
where $\phi(r)$ is the Gaussian basis functions that are scaled by the the state vector. The connectivity kernel can all be decomposed into a set of basis functions
\begin{equation}\label{DefKernelDecomp}
k( \mathbf{r} , \mathbf{r'} ) = \theta^T(\mathbf{r})\psi(\mathbf{r} , \mathbf{r'} )
\end{equation}
Will will assume we know the parametric form of the basis functions, where the parameter $\theta$ is unknown. Each connectivity basis function can be considered a layer in the WC model, representing short range excitation, surround inhibition and long range excitation. Making substitutions we get
\begin{equation}\label{DecompModel}
{\phi ^T}\left( \mathbf{r} \right){x_{t+1}} = \theta^T(\mathbf{r})\int_\Omega  {\psi \left( {{\bf{r}} , {\mathbf{r'}}} \right)f\left( {{\phi ^T}\left( {{\mathbf{r'}}} \right){x_t}} \right)d{\mathbf{r'}}}  - \alpha\phi^T(r)x_t + {\varepsilon _t}\left( {\mathbf{r}} \right).
\end{equation}
To isolate the state we cross multiply by $\phi(r)$ and integrate over $\bf{r}$
\begin{equation}\label{Poo}
\int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}{x_{t+1}}}  = \int_\Omega  {\phi ( \mathbf{r} )\theta^T(\mathbf{r})\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}}  - \alpha\int_\Omega \phi(r)\phi^T(r)d\bf{r} x_t  + \int_\Omega  {\phi \left( {\bf{r}} \right){\varepsilon _t}\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
Now we define
\begin{equation}\label{DefGamma}
\Gamma  = \int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
Substituting equation~\ref{DefGamma} into \ref{Poo} and rearranging gives
\begin{equation}\label{SS Model}
{x_{t+1}} = {\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)\theta^T(\mathbf{r})\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}}-\alpha x_t + {\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right){\varepsilon _t}\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
Under the assumption of spatial homogeneity we can write
\begin{equation}\label{Homogeneous SS Model}
{x_{t+1}} = \theta^T{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)\int_\Omega  {\psi \left( {{\bf{r}} - {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}}-\alpha x_t + {\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right){\varepsilon _t}\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
This provides the final state-space model
\begin{equation}\label{AbbrevSSModel}
x_{t + 1} = \theta^T q(x) -\alpha x + \Gamma ^{ - 1}\int_\Omega  {\phi ( \mathbf{r} ){\varepsilon _t}( \mathbf{r} )d\mathbf{r}}
\end{equation}
with observation equation
\begin{equation}\label{ObservationEquation}
    y_t( k ) = \int_\Omega  m( k,\mathbf{r} )\phi^T( \mathbf{r} ) d\mathbf{r}x_t  + \varepsilon ( k ),
\end{equation}

\section{Estimation of the Nonlinear Homogeneous IDE Neural Field Model}
